---
title: "R for Datascience: Studying Gapminder Dataset"
author: "Anmol Singh"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: spacelab
    highlight: zenburn
    toc: true
    toc_float:
      collapsed: False
      smooth_scroll: False 
    toc_depth: 2
    code_download: true
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
```
```{r}
library(tidyverse)
library(rstatix)
library(MVN)
library(plotly)
library(DT)
library(kableExtra)
```
 

## Question 1:
### Read in the `gapminder_clean.csv` data as a `tibble` using `read_csv`.

```{r}
gapminder <- read_csv("gapminder_clean.csv", col_types = '_') 
datatable(gapminder, options = list(scrollX = TRUE))
``` 
##### We can see that the data contains various demographics to study and compare the growth of many countries from `1962` to `2007`.  
<br>
  
## Question 2:
### Filter the data to include only rows where `Year` is `1962` and then make a scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap` for the filtered data.  
<br>

##### Here we are using log scale for both axes to provide a detailed insight between `'CO2 emissions (metric tons per capita)'` and `gdpPercap`.

```{r, fig.height = 5, fig.width = 10}
gapminder %>% 
  filter(Year == 1962) %>%
  select(`CO2 emissions (metric tons per capita)`, `gdpPercap`) %>% 
  drop_na() %>% 
  ggplot(aes(x = `gdpPercap`,
             y = `CO2 emissions (metric tons per capita)`)) +
  ggtitle("CO2 emissions (metric tons per capita) vs GDP per capita") +
  xlab("GDP per capita") +
  ylab("CO2 emissions (metric tons per capita)") +
  geom_point() +
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(trans = 'log10')
```

##### As we can see from the scatter plot that there is **positive correlation** between `'CO2 emissions (metric tons per capita)'` and `gdpPercap`.  
<br>

## Question 3:
### On the filtered data, calculate the Pearson correlation of `'CO2 emissions (metric tons per capita)'` and `gdpPercap`. What is the Pearson R value and associated p value?  
<br>

##### First we will use a **`Q-Q Plot`** and statistical tests to see if the data we are using satisfies the [assumptions for Pearson correlation](https://statistics.laerd.com/spss-tutorials/pearsons-product-moment-correlation-using-spss-statistics.php).

```{r, fig.height = 4, fig.width = 8}
gapminder %>% 
  filter(`Year` == 1962) %>%
  drop_na(`CO2 emissions (metric tons per capita)`) %>% 
  ggplot(aes(sample = `CO2 emissions (metric tons per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: CO2 emissions (metric tons per capita), Year = 1962") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 1962) %>%
  drop_na(`gdpPercap`) %>% 
  ggplot(aes(sample = `gdpPercap`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: GDP per Capita, Year = 1962") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")
```
```{r,fig.height = 6, fig.width = 10}
# normality tests: 
year = 1962
multivariant <- gapminder %>% 
  filter(Year == year) %>% 
  dplyr::select("CO2 emissions (metric tons per capita)", "gdpPercap") %>% 
  drop_na() %>% 
  mvn(desc = F, mvnTest = "hz", univariateTest = "SW",
      alpha = 0.05, multivariateOutlierMethod = "adj") 

```

### Normality tests for the Year: `r year`

<br>

#### Shapiro-Wilk Univariate-Normality Test: 
```{r,fig.height = 6, fig.width = 10}
multivariant$univariateNormality %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  column_spec(1:5, color = "white",
              background = "#a62a21")

```
<br>

#### Henze-Zirkler Multivariate-Normality Test:
```{r,fig.height = 6, fig.width = 10}
multivariant$multivariateNormality %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  column_spec(1:4, color = "white",
              background = "#a62a21")
  
``` 
<br>

##### According to the result of [Shapiro-Wilk Univariate Normality test and Henze-Zirkler Multivariate Normality test](https://cran.r-project.org/web/packages/MVN/MVN.pdf) and the `Q-Q Plots`, there is enough evidence to show that the data is **neither univariate nor multivariate** since both **`MVN` and `Normality` is ``r multivariant$multivariateNormality$MVN``** for the tests. Also, we can see in the `Adjusted Chi-Square Q-Q Plot` based on *Mahalanobis distance* that there are many outliers in the data. Thus we **failed to satisfy** the assumptions required for Pearson correlation.

##### The assumptions for [Spearman's rank correlation](https://statistics.laerd.com/spss-tutorials/spearmans-rank-order-correlation-using-spss-statistics.php) are satisfied, that is: **variables are in ordinal, interval or ratio scale, two variables represent paired observations and there is monotonic relationship between `"CO2 emissions (metric tons per capita)"` and `"gdpPercap"`**(this is shown in the scatter plot in `Q2`).


```{r}
year = 1962
correlation <- gapminder %>% 
  filter(Year == year)

# Pearson's product-moment correlation
pearson <- cor.test(correlation$`CO2 emissions (metric tons per capita)`,
         correlation$gdpPercap, method = "pearson")

# Spearman's rank correlation
spearman <- cor.test(correlation$`CO2 emissions (metric tons per capita)`,
         correlation$gdpPercap, method = "spearman", exact = TRUE)


```
### Correlation test results for year:`r year` 
#### The Pearson R-value is **``r round(pearson$estimate, digits = 4)``** and associated p-value is **``r sprintf("%.3e", pearson$p.value)``**.
#### The Spearman's Rank correlation statistic is **``r round(spearman$estimate, digits = 4)``** and associated p-value is **`p-value < 2.2e-16`**.  

<br>

##### As we can see for `Spearman's rank correlation` the `rho`(``r round(spearman$estimate, digits = 4)``) > `0`, this means there is a **positive correlation** and `p-value (< 2.2e-16)` < `0.05` meaning that the correlation is **statistically significant**. The `Pearson's correlation` is also **positive**(``r round(pearson$estimate, digits = 4)``) and is **statistically significant**(`p-value`(``r sprintf("%.3e", pearson$p.value)`` < `0.05`).  
<br>  

## Question 4:
### On the unfiltered data, answer "In what year is the correlation between `'CO2 emissions (metric tons per capita)'` and `'gdpPercap'` the strongest?" Filter the dataset to that year for the next step.
<br>

```{r}
years <- unname(unlist(gapminder %>% distinct(Year)))

spearmans_year <- function(year){
 spearman <- gapminder %>% 
    filter(Year == year)
 test <- cor.test(spearman$`CO2 emissions (metric tons per capita)`,
                  spearman$gdpPercap, method = "spearman")
 return(test$estimate)
}

correlation <- lapply(years, spearmans_year)
correlation_df <- as_tibble(cbind(Year = years, Rho = unlist(correlation)))
rownames(correlation_df) <- NULL
correlation_df <- correlation_df[order(correlation_df$Rho, decreasing = TRUE), ]
correlation_df %>% 
  head() %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  row_spec( 1 , bold = T, color = "white", background = "#50C878")


year_high_corr = unlist(correlation_df[1,1])
```
#### The correlation between 'CO2 emissions (metric tons per capita)', and 'gdpPercap' is the strongest in **`r year_high_corr`**.
<br>
```{r}
highest_corr <- gapminder %>%
                  filter(Year == year_high_corr)
datatable(highest_corr, options = list(scrollX = TRUE))

```
<br> 
  
## Question 5:
### Using plotly, create an interactive scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap`, where the point size is determined by `pop` (population) and the color is determined by the `continent`. You can easily convert any `ggplot` plot to a `plotly` plot using the `ggplotly()` command.

```{r, fig.height=6, fig.width=10}
co2_gdp <- highest_corr %>% 
  select(`CO2 emissions (metric tons per capita)`, gdpPercap, 
         continent, pop, `Country Name`) %>% 
  drop_na() %>% 
  ggplot(aes(x = `gdpPercap`,
             y = `CO2 emissions (metric tons per capita)`, 
             size = pop, colour = continent, hoverlabel = `Country Name`)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(.1, 24)) +
  xlab("GDP per capita") +
  ggtitle("CO2 emissions (metric tons per capita) vs GDP per capita") +
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(trans = 'log10')
ggplotly(co2_gdp)
```
<br>  
  
## Question 6:
### What is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`? (Stats test needed)
<br>

##### Let us first see graphically what is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`. Then we will conduct the statistical tests.

```{r, fig.height=6, fig.width=10}

box_plot <- gapminder %>%
  select(`continent`, `Energy use (kg of oil equivalent per capita)`,
         `Country Name`, Year) %>%
  drop_na() %>% 
  ggplot(aes(x = continent,
             y = `Energy use (kg of oil equivalent per capita)`, 
             hovername = `Country Name`,
             frame = Year)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5, colour = "brown") +
  xlab("Continents") +
  ylab("Energy use (kg of oil equivalent per capita)") +
  ggtitle("Energy use (kg of oil equivalent per capita) vs Continents") +
  scale_y_continuous(trans = 'log10', limits = c(9, 15000)) 
ggplotly(box_plot)

```
<br>

##### Before moving forward, let us make sure if the data follows **parametric assumptions**. To check the distribution, we can use **`Quantile-Quantile Plots (QQ-plots)`**.

```{r}

gapminder %>% 
  filter(`Year` == 2007, continent == "Asia") %>%
  drop_na(`Energy use (kg of oil equivalent per capita)`) %>% 
  ggplot(aes(sample = `Energy use (kg of oil equivalent per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Energy use (kg of oil equivalent per capita),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 2007, continent == "Europe") %>%
  drop_na(`Energy use (kg of oil equivalent per capita)`) %>% 
  ggplot(aes(sample = `Energy use (kg of oil equivalent per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Energy use (kg of oil equivalent per capita), 
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

```
<br>  

##### As we can see the data is **not normally distributed** for every continent across the years, thus we **cannot use parametric tests**(example: one-way ANOVA). So, to show the statistical relation between `continent` and `'Energy use (kg of oil equivalent per capita)'` we will use the [Kruskal-Wallis Test](https://rpkgs.datanovia.com/rstatix/reference/kruskal_test.html) which is a **non-parametric test**.
<br>

#### **Kruskal-Wallis Test results**:
```{r}
kruskal <- function(year){
  krusk_wallis <- gapminder %>% 
    filter(Year == year) %>%
    select(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    drop_na(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    kruskal_test(`Energy use (kg of oil equivalent per capita)` ~ continent)
  krusk_wallis <- cbind(Year = year, krusk_wallis,
                        signif_p_val = krusk_wallis$p < 0.05) %>% 
    select(-.y.)
}
kruskal_wall <- lapply(years, kruskal)
kruskal_wall <- as_tibble(do.call(rbind, kruskal_wall))
kruskal_wall <- rename(kruskal_wall, degree_of_freedom = df,
                       p_value = p,  sample_count = n)
kruskal_wall %>% 
  select(Year, sample_count, statistic,  p_value, signif_p_val) %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  column_spec(1:5, color = "white",
              background = ifelse(kruskal_wall$p_value > 0.05,"#a62a21", "#50C878"))
```

##### It can be observed that excluding the years `1962` and `1967`, the `p-value` < `0.05`. In conclusion, there is a **significant relationship** between **`continents`** and **`Energy use (kg of oil equivalent per capita)`** for years **`1972`** to **`2007`**. It is suggested to do a **post hoc test** to determine detailed relationship between groups. We will conduct [Dunn's Test](https://rdrr.io/cran/rstatix/man/dunn_test.html) for the same using the `Bonferonni correction`(multiple testing correction) for the `p-values`.
<br>

#### **Dunn's Test results:** 
```{r}
dunn <- function(year){
  dunns <- gapminder %>% 
    filter(Year == year) %>%
    select(`Energy use (kg of oil equivalent per capita)`, continent) %>%
    drop_na(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    dunn_test(`Energy use (kg of oil equivalent per capita)` ~ continent, 
              p.adjust.method = "bonferroni")
  dunn <- cbind(Year = year, dunns) %>% 
    select(-.y.)
}

dunns_test <- lapply(years[years > 1967], dunn) 
dunns_test <- as_tibble(do.call(rbind, dunns_test))
dunns_test <- dunns_test %>% 
  mutate(p.adj.signif = (p.adj.signif != 'ns'))
dunns_test <- rename(dunns_test, p_value = p, p_adj = p.adj, 
                     signif_p_adj = p.adj.signif)

dunns_test %>% 
  select(Year, group1, group2, statistic, p_adj, signif_p_adj) %>%
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  column_spec(1:6, color = "white",
              background = ifelse(dunns_test$p_adj > 0.05,"#a62a21", "#50C878")) %>% 
  scroll_box( height = "500px", fixed_thead = TRUE)


```
#### As we can see from the table there is a **significant relationship** between many `continents`. 
* **Example**:  
    + **`Africa` and `Europe` in `1972`** 
    + **`Americas` and `Europe` in `1982`** 
<br>

## Question 7:
### Is there a significant difference between `Europe` and `Asia` with respect to `'Imports of goods and services (% of GDP)'` in the years after `1990`? (Stats test needed)
<br>

##### To check the distribution of `'Imports of goods and services (% of GDP)'` for `Europe` and `Asia`, we can use **`Quantile-Quantile Plots (Q-Q plots)`**.

```{r}
gapminder %>% 
  filter(`Year` == 1992, continent == "Asia") %>%
  drop_na(`Imports of goods and services (% of GDP)`) %>% 
  ggplot(aes(sample = `Imports of goods and services (% of GDP)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Imports of goods and services (% of GDP),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 1992, continent == "Europe") %>%
  drop_na(`Imports of goods and services (% of GDP)`) %>% 
  ggplot(aes(sample = `Imports of goods and services (% of GDP)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Imports of goods and services (% of GDP),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

```

##### As we can observe from the **`Q-Q plots`** that there are **outliers** in the data and the data is **not normally distributed**. Thus to check if there is a significant difference between `Europe` and `Asia` with respect to `'Imports of goods and services (% of GDP)'` in the years after `1990`. We will be using [Wilcoxon Rank Sum test](https://rpkgs.datanovia.com/rstatix/reference/wilcox_test.html).
<br> 

#### **Wilcoxon Rank Sum test results:**
```{r}
wilcox <- function(year){
 wilcoxs <- gapminder %>% 
   filter(Year == year, continent == 'Asia' | continent == 'Europe') %>%
   select(`Imports of goods and services (% of GDP)`, continent) %>% 
   drop_na(`Imports of goods and services (% of GDP)`, continent) %>% 
   wilcox_test(`Imports of goods and services (% of GDP)` ~ continent)
 wilcoxs <- cbind(Year = year, wilcoxs,  
               signif_p_val = wilcoxs$p < 0.05) %>% 
   select(-.y.)
}
wilcoxs_rank <- lapply(years[years > 1990], wilcox)
wilcoxs_rank <- as_tibble(do.call(rbind, wilcoxs_rank))
wilcoxs_rank %>% 
  select(Year, group1, group2, statistic, p, signif_p_val) %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  column_spec(1:6, color = "white",
              background = ifelse(wilcoxs_rank$p > 0.05,"#a62a21", "#50C878"))

```

##### As we can see in all the years after `1990` the **`p > 0.05`**, this means that there is **no significant difference** between **`Europe`** and **`Asia`** with respect to **`'Imports of goods and services (% of GDP)'`** in the years after `1990`.
<br>  

## Question 8:
### What is the country (or countries) that has the highest `'Population density (people per sq. km of land area)'` across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)

```{r}
gapminder %>% 
  select(`Population density (people per sq. km of land area)`,
         `Country Name`, `Year`) %>% 
  drop_na() %>% 
  group_by(`Year`) %>% 
  mutate(avg_pop_rank = 
          rank(desc(`Population density (people per sq. km of land area)`))) %>% 
  ungroup(`Year`) %>% 
  select(`Country Name`, avg_pop_rank) %>% 
  group_by(`Country Name`) %>% 
  mutate(avg_pop_rank =  mean(avg_pop_rank)) %>%  
  ungroup(`Country Name`) %>% 
  unique() %>% 
  arrange(avg_pop_rank) %>%
  head() %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  row_spec( 1:2 , bold = T, color = "white", background = "#50C878")
  
```

##### As we can see **`Macao SAR, China`** and **`Monaco`** both with **`avg_pop_rank`** of **`1.5`** had the highest `'Population density (people per sq. km of land area)'` across `1962` to `2007`.
<br>  

## Question 9:
### What country (or countries) has shown the greatest increase in `'Life expectancy at birth, total (years)'` since `1962`?

```{r}

gapminder %>% 
  select(`Life expectancy at birth, total (years)`,
         `Country Name`, `Year`) %>% 
  drop_na() %>% 
  group_by(`Country Name`) %>% 
  filter(Year == 1962 | Year == 2007) %>% 
  mutate(diff_life_expectancy = `Life expectancy at birth, total (years)` 
         - lag( `Life expectancy at birth, total (years)`)) %>% 
  select(`Country Name`, diff_life_expectancy) %>% 
  drop_na() %>% 
  arrange(-diff_life_expectancy) %>% 
  head() %>% 
  kbl() %>% 
  kable_paper(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) %>% 
  kable_styling(fixed_thead = T) %>% 
  row_spec( 1 , bold = T, color = "white", background = "#50C878")

  
```

##### We can see that **`Maldives`** had the greatest increase in `'Life expectancy at birth, total (years)'` since `1962` with an increase of **`36.916`** from `1962` to `2007`.  
