---
title: "R for Datascience: Studying Gapminder Dataset"
author: "Anmol Singh"
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE) 
```
  
**Packages used:**
```{r}
library(tidyverse)
library(rstatix)
library(MVN)
library(plotly)
```

#### Q1: Read in the `gapminder_clean.csv` data as a `tibble` using `read_csv`.

```{r}
gapminder <- read_csv("gapminder_clean.csv", col_types = '_') 
gapminder %>% 
  print(n = 5)
``` 
We can see that the data contains various demographics to study and compare the growth of many countries from `1962` to `2007`.

#### Q2: Filter the data to include only rows where `Year` is `1962` and then make a scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap` for the filtered data.  

Here we are using log scale for both axes to provide a detailed insight between `'CO2 emissions (metric tons per capita)'` and `gdpPercap`.

```{r, fig.height = 5, fig.width = 10}
gapminder %>% 
  filter(Year == 1962) %>%
  select(`CO2 emissions (metric tons per capita)`, `gdpPercap`) %>% 
  drop_na() %>% 
  ggplot(aes(x = `gdpPercap`,
             y = `CO2 emissions (metric tons per capita)`)) +
  ggtitle("CO2 emissions (metric tons per capita) vs GDP per capita") +
  xlab("GDP per capita") +
  ylab("CO2 emissions (metric tons per capita)") +
  geom_point() +
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(trans = 'log10')
```

As we can see from the scatter plot that there is positive correlation between `'CO2 emissions (metric tons per capita)'` and `gdpPercap`.

#### Q3: On the filtered data, calculate the Pearson correlation of `'CO2 emissions (metric tons per capita)'` and `gdpPercap`. What is the Pearson R value and associated p value?  

First we will use a `Q-Q Plot` and statistical tests to see if the data we are using satisfies the [assumptions for Pearson correlation](https://statistics.laerd.com/spss-tutorials/pearsons-product-moment-correlation-using-spss-statistics.php).

```{r, fig.height = 4, fig.width = 8}
gapminder %>% 
  filter(`Year` == 1962) %>%
  drop_na(`CO2 emissions (metric tons per capita)`) %>% 
  ggplot(aes(sample = `CO2 emissions (metric tons per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: CO2 emissions (metric tons per capita), Year = 1962") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 1962) %>%
  drop_na(`gdpPercap`) %>% 
  ggplot(aes(sample = `gdpPercap`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: GDP per Capita, Year = 1962") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")
```
```{r,fig.height = 6, fig.width = 10}
# normality tests: 
gapminder %>% 
  filter(Year == 1962) %>% 
  dplyr::select("CO2 emissions (metric tons per capita)", "gdpPercap") %>% 
  drop_na() %>% 
  mvn(desc = F, mvnTest = "hz", univariateTest = "SW",
      alpha = 0.05, multivariateOutlierMethod = "adj")
```

According to the result of [Shapiro-Wilk Univariate Normality test and Henze-Zirkler Multivariate Normality test](https://cran.r-project.org/web/packages/MVN/MVN.pdf) and the `Q-Q Plots`, there is enough evidence to show that the data is neither univariate nor multivariate since both `MVN` and `Normality` is `NO` for the tests. Also, we can see in the `Adjusted Chi-Square Q-Q Plot` based on Mahalanobis distance that there are many outliers in the data. Thus we failed to satisfy the assumptions required for Pearson correlation.

As the question explicitly asks for Pearson's correlation, the results of both Pearson's correlation and Spearman's rank correlation are included.

The assumptions for [Spearman's rank correlation](https://statistics.laerd.com/spss-tutorials/spearmans-rank-order-correlation-using-spss-statistics.php) are satisfied, that is: variables are in ordinal, interval or ratio scale, two variables represent paired observations and there is monotonic relationship between `"CO2 emissions (metric tons per capita)"` and `"gdpPercap"`(this is shown in the scatter plot in `Q2`).

```{r}
correlation <- gapminder %>% 
  filter(Year == 1962)
print('Year: 1962')
# Pearson's product-moment correlation
cor.test(correlation$`CO2 emissions (metric tons per capita)`,
         correlation$gdpPercap, method = "pearson")
# Spearman's rank correlation
cor.test(correlation$`CO2 emissions (metric tons per capita)`,
         correlation$gdpPercap, method = "spearman")
```

As we can see for `Spearman's rank correlation` the `rho`(`0.853`) > `0`, this means there is a positive correlation and `p-value`(`< 2.2e-16`) < `0.05` meaning that the correlation is statistically significant. The `Pearson's correlation` is also positive(`cor` = `0.926`) and is statistically significant(`p-value`(`2.2e-16`) < `0.05`).

#### Q4: On the unfiltered data, answer "In what year is the correlation between `'CO2 emissions (metric tons per capita)'` and `'gdpPercap'` the strongest?" Filter the dataset to that year for the next step.

```{r}
years <- unname(unlist(gapminder %>% distinct(Year)))

spearmans_year <- function(year){
 spearman <- gapminder %>% 
    filter(Year == year)
 test <- cor.test(spearman$`CO2 emissions (metric tons per capita)`,
                  spearman$gdpPercap, method = "spearman")
 return(test$estimate)
}

correlation <- lapply(years, spearmans_year)
correlation_df <- as_tibble(cbind(Year = years, Rho = unlist(correlation)))
rownames(correlation_df) <- NULL
correlation_df <- correlation_df[order(correlation_df$Rho, decreasing = TRUE), ]
correlation_df %>% 
  print(n = 5)
year_high_corr = unlist(correlation_df[1,1])
print(paste("The correlation between 'CO2 emissions (metric tons per capita)'",
            "and 'gdpPercap' is the strongest in", year_high_corr))

highest_corr <- gapminder %>%
                  filter(Year == year_high_corr)
highest_corr %>% 
  print(n = 5)

```

#### Q5:Using plotly, create an interactive scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap`, where the point size is determined by `pop` (population) and the color is determined by the `continent`. You can easily convert any `ggplot` plot to a `plotly` plot using the `ggplotly()` command.

```{r, fig.height=6, fig.width=10}
co2_gdp <- highest_corr %>% 
  select(`CO2 emissions (metric tons per capita)`, gdpPercap, 
         continent, pop, `Country Name`) %>% 
  drop_na() %>% 
  ggplot(aes(x = `gdpPercap`,
             y = `CO2 emissions (metric tons per capita)`, 
             size = pop, colour = continent, hoverlabel = `Country Name`)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(.1, 24)) +
  xlab("GDP per capita") +
  ggtitle("CO2 emissions (metric tons per capita) vs GDP per capita") +
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(trans = 'log10')
ggplotly(co2_gdp)
```

#### Q6: What is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`? (Stats test needed)

Let us first see graphically what is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`. Then we will conduct the statistical tests.

```{r, fig.height=6, fig.width=10}

box_plot <- gapminder %>%
  select(`continent`, `Energy use (kg of oil equivalent per capita)`,
         `Country Name`, Year) %>%
  drop_na() %>% 
  ggplot(aes(x = continent,
             y = `Energy use (kg of oil equivalent per capita)`, 
             hovername = `Country Name`,
             frame = Year)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5, colour = "brown") +
  xlab("Continents") +
  ylab("Energy use (kg of oil equivalent per capita)") +
  ggtitle("Energy use (kg of oil equivalent per capita) vs Continents") +
  scale_y_continuous(trans = 'log10', limits = c(9, 15000)) 
ggplotly(box_plot)

```

Before moving forward, let us make sure if the data follows parametric assumptions. To check the distribution, we can use `Quantile-Quantile Plots (QQ-plots)`.

```{r}

gapminder %>% 
  filter(`Year` == 2007, continent == "Asia") %>%
  drop_na(`Energy use (kg of oil equivalent per capita)`) %>% 
  ggplot(aes(sample = `Energy use (kg of oil equivalent per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Energy use (kg of oil equivalent per capita),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 2007, continent == "Europe") %>%
  drop_na(`Energy use (kg of oil equivalent per capita)`) %>% 
  ggplot(aes(sample = `Energy use (kg of oil equivalent per capita)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Energy use (kg of oil equivalent per capita), 
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

```

As we can see the data is not normally distributed for every continent across the years, thus we cannot use parametric tests(example: one-way ANOVA). So, to show the statistical relation between `continent` and `'Energy use (kg of oil equivalent per capita)'` we will use the [Kruskal-Wallis Test](https://rpkgs.datanovia.com/rstatix/reference/kruskal_test.html) which is a non-parametric test.

```{r}
kruskal <- function(year){
  krusk_wallis <- gapminder %>% 
    filter(Year == year) %>%
    select(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    drop_na(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    kruskal_test(`Energy use (kg of oil equivalent per capita)` ~ continent)
  krusk_wallis <- cbind(Year = year, krusk_wallis,
                        signif_p_val = krusk_wallis$p < 0.05) %>% 
    select(-.y.)
}
kruskal_wall <- lapply(years, kruskal)
kruskal_wall <- as_tibble(do.call(rbind, kruskal_wall))
kruskal_wall <- rename(kruskal_wall, degree_of_freedom = df,
                       p_value = p,  sample_count = n)
kruskal_wall
```

It can be observed that excluding the years `1962` and `1967`, the `p-value` < `0.05`. Also, `degree_of_freedom` = `3` for `1962` and `1967` as `Africa` has only `NaN` values for year `1962` and `1967`. In conclusion, there is a significant relationship between `continents` and `Energy use (kg of oil equivalent per capita)` for years `1972` to `2007`. It is suggested to do a post hoc test to determine detailed relationship between groups. We will conduct [Dunn's Test](https://rdrr.io/cran/rstatix/man/dunn_test.html) for the same using the `Bonferonni correction`(multiple testing correction) for the `p-values`.

```{r}
dunn <- function(year){
  dunns <- gapminder %>% 
    filter(Year == year) %>%
    select(`Energy use (kg of oil equivalent per capita)`, continent) %>%
    drop_na(`Energy use (kg of oil equivalent per capita)`, continent) %>% 
    dunn_test(`Energy use (kg of oil equivalent per capita)` ~ continent, 
              p.adjust.method = "bonferroni")
  dunn <- cbind(Year = year, dunns) %>% 
    select(-.y.)
}

dunns_test <- lapply(years[years > 1967], dunn) 
dunns_test <- as_tibble(do.call(rbind, dunns_test))
dunns_test <- dunns_test %>% 
  mutate(p.adj.signif = (p.adj.signif != 'ns'))
dunns_test <- rename(dunns_test, p_value = p, p_adj = p.adj, 
                     signif_p_adj = p.adj.signif)
dunns_test %>% 
  print(n = 5)

# But we are mainly interested where there is a statistically significant relationship between groups
dunns_test %>% 
  filter(signif_p_adj == T) %>% 
  print(n = 5)

```
As we can clearly see that for groups where `signif_p_adj` is `TRUE`, there is a significant relationship between the groups(`continent`). 
Example: `Africa` and `Europe` in `1972`, `Americas` and `Europe` in `1982`.

### Q7: Is there a significant difference between `Europe` and `Asia` with respect to `'Imports of goods and services (% of GDP)'` in the years after `1990`? (Stats test needed)

To check the distribution of `'Imports of goods and services (% of GDP)'` for `Europe` and `Asia`, we can use `Quantile-Quantile Plots (Q-Q plots)`.

```{r}
gapminder %>% 
  filter(`Year` == 1992, continent == "Asia") %>%
  drop_na(`Imports of goods and services (% of GDP)`) %>% 
  ggplot(aes(sample = `Imports of goods and services (% of GDP)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Imports of goods and services (% of GDP),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

gapminder %>% 
  filter(`Year` == 1992, continent == "Europe") %>%
  drop_na(`Imports of goods and services (% of GDP)`) %>% 
  ggplot(aes(sample = `Imports of goods and services (% of GDP)`)) +
  stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot: Asia: Imports of goods and services (% of GDP),
          Year = 2007") +
  xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

```

As we can observe from the `Q-Q plots` that there are outliers in the data and the data is not normally distributed. Thus to check if there is a significant difference between `Europe` and `Asia` with respect to `'Imports of goods and services (% of GDP)'` in the years after `1990`. We will be using [Wilcoxon Rank Sum test](https://rpkgs.datanovia.com/rstatix/reference/wilcox_test.html).

```{r}
wilcox <- function(year){
 wilcoxs <- gapminder %>% 
   filter(Year == year, continent == 'Asia' | continent == 'Europe') %>%
   select(`Imports of goods and services (% of GDP)`, continent) %>% 
   drop_na(`Imports of goods and services (% of GDP)`, continent) %>% 
   wilcox_test(`Imports of goods and services (% of GDP)` ~ continent)
 wilcoxs <- cbind(Year = year, wilcoxs,  
               signif_p_val = wilcoxs$p < 0.05) %>% 
   select(-.y.)
}
wilcoxs_rank <- lapply(years[years > 1990], wilcox)
wilcoxs_rank <- as_tibble(do.call(rbind, wilcoxs_rank))
wilcoxs_rank

```

As we can see in all the years after `1990` the `p` > `0.05`, this means that there is no significant difference between `Europe` and `Asia` with respect to `'Imports of goods and services (% of GDP)'` in the years after `1990`.

### Q8: What is the country (or countries) that has the highest `'Population density (people per sq. km of land area)'` across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)

```{r}
gapminder %>% 
  select(`Population density (people per sq. km of land area)`,
         `Country Name`, `Year`) %>% 
  drop_na() %>% 
  group_by(`Year`) %>% 
  mutate(avg_pop_rank = 
          rank(desc(`Population density (people per sq. km of land area)`))) %>% 
  ungroup(`Year`) %>% 
  select(`Country Name`, avg_pop_rank) %>% 
  group_by(`Country Name`) %>% 
  mutate(avg_pop_rank =  mean(avg_pop_rank)) %>%  
  ungroup(`Country Name`) %>% 
  unique() %>% 
  arrange(avg_pop_rank) %>% 
  print(n = 5)
  
```

As we can see `Macao SAR, China` and `Monaco` both with `avg_pop_rank` of `1.5` had the highest `'Population density (people per sq. km of land area)'` across `1962` to `2007`.

### Q9: What country (or countries) has shown the greatest increase in 'Life expectancy at birth, total (years)' since 1962?

```{r}

gapminder %>% 
  select(`Life expectancy at birth, total (years)`,
         `Country Name`, `Year`) %>% 
  drop_na() %>% 
  group_by(`Country Name`) %>% 
  filter(Year == 1962 | Year == 2007) %>% 
  mutate(diff_life_expectancy = `Life expectancy at birth, total (years)` 
         - lag( `Life expectancy at birth, total (years)`)) %>% 
  select(`Country Name`, diff_life_expectancy) %>% 
  drop_na() %>% 
  arrange(-diff_life_expectancy) %>% 
  print(n = 5)
  
```

We can see that `Maldives` had the greatest increase in `'Life expectancy at birth, total (years)'` since `1962` with an increase of `36.916` from `1962` to `2007`.